Statically Typed Web Programming using Scala

### 2012-12-20 papers/ecoop2013/repairing.tex
Some words on the asynchronous programming DSL implementation

Asynchronous values can be transformed using the \code{map} and \code{flatMap} methods, turning them into first-class
citizens: functions can take them as parameters and return them.

\begin{figure}
\begin{multicols}{2}
\begin{lstlisting}[caption=Parallel computations in Scala,label=async-parallel-1]
val fooAsync = Ajax.get("/foo")
val barAsync = Ajax.get("/bar")
for {
  foo <- fooAsync
  bar <- barAsync
} println(foo + bar)
\end{lstlisting}
\vfill
\columnbreak
\begin{lstlisting}[language=JavaScript,caption=Generated JavaScript code,label=async-parallel-2]
var x1 = { cs: [] };
AjaxGet("/foo", x1);
var x2 = { cs: [] };
AjaxGet("/bar", x2);
foreach(x1, function (foo) {
 foreach(x2, function (bar) {
  var x3 = foo + bar;
  console.log(x3);
 });
});
\end{lstlisting}
\end{multicols}
\end{figure}

Listing \ref{async-parallel-1} defines two asynchronous computations running in parallel and listing
\ref{async-parallel-2} shows the generated JavaScript code (we renamed some identifiers for the sake of readability).
The \code{x1} and \code{x2} values encode the asynchronous results of the Ajax requests, they contain an array of
callbacks (\code{cs}, initially empty) to call when the result is available. The \code{AjaxGet} function code has
been omitted, it creates a \code{XMLHttpRequest} object that sends a HTTP request and completes the value of its
last parameter with the response, when received. Completing an asynchronous value means defining a \code{v} field
on it containing the actual value, and calling all pending callbacks that are in the \code{cs} field. The
\code{foreach} code has also been omitted, it registers a callback on an asynchronous value or calls the callback
directly with the value if this one is already available.

@@

>>>>

Asynchronous values can be used and transformed using the \code{foreach}, \code{map} and \code{flatMap} methods, turning them into first-class
citizens: functions can take them as parameters and return them. Consequently, it is easier to write concurrent code:
listing \ref{async-parallel-1} shows how to run two parallel asynchronous computations and to do something with their
results when both are available. Writing equivalent code in pure JavaScript is tedious because it requires that the
callback passed to each asynchronous computation checks if the other has been completed before, and to store the
value, when availalbe, in a shared place (so both callbacks can get the other’s value).

\begin{figure}
\begin{lstlisting}[caption=Parallel computations in Scala,label=async-parallel-1]
val fooAsync = Ajax.get("/foo")
val barAsync = Ajax.get("/bar")
for {
  foo <- fooAsync
  bar <- barAsync
} println(foo + bar)
\end{lstlisting}
\end{figure}

@@

\subsubsection{Taking advantage of the three transformation steps}

The asynchronous programming DSL makes use of a combination of the three transformation steps described above.
Asynchronous values are reified in values with a \code{put} and a \code{get} operations. The \code{put} operation
sets the value of the asynchronous value and the \code{get} operation uses the asynchronous value when available.
All other operations (\code{foreach}, \code{map} and \code{flatMap}) are implemented in terms of these two primitives.

\begin{figure}
\begin{lstlisting}[language=JavaScript,caption={Generated JavaScript code for listing \protect\ref{async-parallel-1}},label=async-parallel-2]
var x1 = { v: null, cs: [] };
AjaxGet("/foo", x1);
var x2 = { v: null, cs: [] };
AjaxGet("/bar", x2);
var Get = function (f, callback) {
  if (f.v === null) f.cs.push(callback)
  else callback(f.v)
};
Get(x1, function (foo) {
 Get(x2, function (bar) {
  var x3 = foo + bar;
  console.log(x3);
 });
});
\end{lstlisting}
\end{figure}

Listing \ref{async-parallel-2} shows the generated JavaScript code for the Scala listing \ref{async-parallel-1} (we
renamed some identifiers for the sake of readability). The \code{x1} and \code{x2} values reify the asynchronous
results of the Ajax requests, they contain a field \code{v} holding the result value (initialized to \code{null}) and an
array of callbacks \code{cs} (initially empty) to call when the result is available. The \code{AjaxGet} function code
has been omitted, it creates a \code{XMLHttpRequest} object that sends a HTTP request and completes the value of its
last parameter with the response, when received, using the \code{put} operation. The implementation of \code{put} sets
the \code{v} field to the value and calls all the pending \code{cs} callbacks. The \code{Get} function shows the
implementation of the \code{get} primitive. It takes two parameters, an asynchronous value and a callback. If the value
is not yet available it adds the callback on the pending callback list, otherwise it calls it with the value as
parameter.

The transformation process of code using the asynchronous programming DSL can be summarized as follows. First, the
compilation step desugars the \code{for} notation to calls to \code{foreach}, \code{map} and \code{flatMap}. These
methods are implemented in terms of the \code{put} and \code{get} primitives, so they evaluate to an intermediate
representation containing \code{Put} and \code{Get} nodes. Finally, these nodes are translated into the code actually
implementing the asynchronous values reification.
